Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-06 rnni_convexity.tex
abstract + paper structure

\begin{abstract}
In this paper we study the graph of ranked phylogenetic trees where the adjacency relation is given by a local rearrangement of the tree structure.
Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of the optimisation.
We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
Specifically, we introduce some algorithms for computing paths in the $\rnni$ graph and use them to establish properties such as diameter and radius of this graph.

We also study the problem of computing distances between trees in this graph.
Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we specifically focus on the properties that have already been established for $\nni$.
Surprisingly, our results suggest that the complexity of computing distances in the two graphs is likely to be different.
%TODO
% Alex: The proof for $\np$ hardness of $\nni$ uses trees that have two caterpillar trees as subtrees (See Fig5).
% I'm not sure whether it's known that computing distances just between caterpillar trees is not polynomial in $\nni$.
% Do I need to be more specific about this here?
% And shall I use the term 'caterpillar trees' here?
We prove that within a particular subset of trees distances can be computed in polynomial time in $\rnni$ while this is not known to be true in $\nni$.
\end{abstract}

@@

The rest of this paper is structured as follows.
Within the next section we introduce three algorithms for exploring the $\rnni$ graph.
We use these algorithms in Section~\ref{section:geometry} for proving some geometrical properties of this graph, such as diameter and radius.
Specifically, we will prove in Section~\ref{section:caterpillar_convex} that for a subset of $\rnni$ trees, distances can be computed in polynomial time.
This already suggests that there is a difference in complexity of computing distances between the $\rnni$ graph and the classical $\nni$ graph.
\todo{Alex: Is it OK to use 'shortest $\rnni$ paths' here?}
We will discuss this further in Section~\ref{section:cluster_theorem} where we disprove a conjecture raised in \autocite{Gavryushkin2018-ol} and provide an alternative version, which makes a claim about the similarity of trees on shortest $\rnni$ paths.

>>>>

\begin{abstract}
In this paper we study the graph of ranked phylogenetic trees where the adjacency relation is given by a local rearrangement of the tree structure.
Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of the optimisation.
We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
For example, we find exact estimates of the radius and diameter of the space.
Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we compare geometric and algorithmic properties of the two graphs.
Surprisingly, we discover that both geometric and algorithmic properties of $\rnni$ and $\nni$ are quite different.
For example, we establish convexity of certain natural subspaces in $\rnni$, which is not convex is $\nni$.
Furthermore, our results suggest that the complexity of computing distances in the two graphs is different.
\end{abstract}

@@

The rest of this paper is organised as follows.
In Section~\ref{section:algorithms} we introduce three algorithms for exploring the $\rnni$ graph.
We use these algorithms in Section~\ref{section:geometry} to establish several geometrical properties of this graph, such as its diameter and radius.
Furthermore, we will prove in Section~\ref{section:caterpillar_convex} that distances can be computed in polynomial time in a subspace of the $\rnni$ space.
This result suggests that the complexity of computing distances in $\rnni$ is different from that in the classical $\nni$ space.
We will discuss this further in Section~\ref{section:cluster_theorem}, where we disprove a conjecture of \textcite{Gavryushkin2018-ol} and suggest its weaker version a plausible alternative.
